---
engine: knitr
editor_options: 
  chunk_output_type: console
---

# Sample selection

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(dev = "svg",
                      comment = "#>",
                      dev.args = list(bg = "transparent"))

library(tidyverse)
library(readxl)
library(here)
library(DT)
source(here("R", "theme_kh.R"))
set.seed(42)

clrs <- c(Recruited = clr1,
          NonRecruited = clr2,
          BeachDead = clr3)
```

Samples with issues, to avoid for sample selection.

```{r}
troubblemakers <- c(
   "AGP00150", # 2_explore..., l 23, duplicated tissue id
   "AGP08134", # 2_explore..., l 65 ff potential sample mixup
   "AGP08139", #                       continued
   "AGP18084"  # 2_explore..., l 69, mum genetic mismatch
   # REMEMBER TO ALSO EXCLUDE THE SAMPLES THAT HAVE ALREADY BEEN RESEQED
)
```


Read in Annekes cleaned sample list.
Then, give an overview of the available samples over the years, stratified by recruitment success and SNP genotype availability.

```{r}
#| figure-width: 7
#| fig-height: 5
data <- readxl::read_xlsx(here("../import_anneke/data/processed/Unique_filtered_dataset_2022.xlsx")) |> 
  dplyr::select(PupTissueID, uniqueID_pup, PupBirthyear, status, PupSex, pup_SNPed, Beach) |> 
  filter(Beach == "SSB",
         PupSex == "F",
         !is.na(PupBirthyear),
         !(PupTissueID %in% troubblemakers))

# data |> 
#   group_by(uniqueID_pup) |> 
#   count() |> arrange(-n) 
#  # group_by(n) |>
#  # count()

data |> 
  filter(!duplicated(uniqueID_pup)) |> 
  group_by(PupBirthyear, status, pup_SNPed) |> 
  count() |> 
  ggplot() +
  # geom_hline(yintercept = 2, linetype = 3 ) +
  geom_bar(aes(x = PupBirthyear, y = n, fill = status),
           stat = "identity",
           position = position_dodge2(width = 0.9, preserve = "single"),
           color = "transparent") +
  facet_grid(pup_SNPed ~ .,
             labeller = label_both,
             switch = "y") +
  scale_fill_manual(values = clrs) +
  theme(legend.position = "bottom",
        axis.line.x = element_line(size = .2),
        legend.key.height = unit(2, "pt"))
```

For each year, two samples are randomly chosen per recruitment status.
Try to use only SNPed samples, if necessary  fill up to two samples where possible (if only one sample is available, pick this one).

```{r}
# best_concentrations <- readxl::read_xlsx(here("../import_anneke/data/processed/concentrations_remaining_recruits.xlsx")) |> 
#   bind_rows(readxl::read_xlsx(here("../import_anneke/data/processed/concentrations_remaining_nonrecruits_with_SNPed_mums.xlsx"))) |> 
#   dplyr::select(uniqueID_pup, SampleID:Gel) |> 
#   arrange(uniqueID_pup, -Picogreen_con) |> 
#   filter(!duplicated(uniqueID_pup))

# sample_to_id <- readxl::read_xlsx(here("../import_anneke/data/processed/Trains_metadata_2022.xlsx")) |> 
#    dplyr::select(SampleID = PupTissueID, uniqueID_pup)

data_con <- readxl::read_xlsx(here("../import_anneke/data/processed/Agaz_DNA_extraction_info.xlsx")) |> 
   rename(PupTissueID = SampleID)
 
data_with_con <- data |> 
  mutate(fitness_class = if_else(status %in% c("BeachDead", "NonRecruited"),
                                 "NonRecruited", status)) |> 
  filter(!duplicated(uniqueID_pup),
         PupSex == "F",
         fitness_class %in% c("Recruited", "NonRecruited") ) |>
  left_join(data_con)

data_top_con <- data_with_con |> 
  group_by(PupBirthyear, status, pup_SNPed) |> 
  mutate(con_where_possible = if_else(is.na(Picogreen_con) | Picogreen_con < 45,
                                      runif(n = length(Picogreen_con), -150, -100),
                                      Picogreen_con)) |> 
  arrange(PupBirthyear, status, pup_SNPed, -con_where_possible) |>
  mutate(con_rank = row_number(),
         con_n = n()) |> 
  # filter(con_rank < 4) |> 
  ungroup()
```


```{r}
data_selection <- data_top_con |> 
  mutate(fitness_class = if_else(status %in% c("BeachDead", "NonRecruited"),
                                 "NonRecruited", status)) |> 
  filter(!duplicated(uniqueID_pup),
         PupSex == "F",
         fitness_class %in% c("Recruited", "NonRecruited") ) |> 
  group_by(PupBirthyear, status, pup_SNPed) |> 
  slice_sample(n = 1, weight_by = scales::rescale(con_where_possible)) |> 
  # mutate(selector = sample(1:n())) |> # random selection order within year x fitness_class
  # filter(selector <= 2) |>            # slice_sample(n = 2) cant subset for years with only one sample 
  ungroup() |> 
  arrange(PupBirthyear, status, pup_SNPed) |> 
  group_by(PupBirthyear, status) |> 
  mutate(priority = row_number()) |>
  ungroup() |> 
  filter(priority <= 1) |>             # if possible, use SNPed samples, fill up if necessary
  mutate(uniqueID_pup = if_else(is.na(uniqueID_pup),   # the sample from 1977 is IDed as <NA>
                                "ID_00", uniqueID_pup)) |> # change to ID_00
  mutate(y = c(Recruited = .5,
               NonRecruited = -.5,
              BeachDead = -1.5)[status])                      
```

Looking up available alternative tissue samples for the selected individuals.

```{r}
# Collect alternative tissue IDs
uniqueIDs <- openxlsx::read.xlsx(here("../import_anneke/data/processed/all_msat_genotypes_uniqueID.xlsx")) %>% 
  dplyr::select(dummyID, PlateNumber, uniqueID) |> 
  filter(!dummyID %in% c("AGP08134", "AGP18084")) |>
  dplyr::select(-PlateNumber) |> 
  unique()

# Merge IDs with list of remaining individuals
tissue_options <- data_selection |> 
  left_join(uniqueIDs, by = c("uniqueID_pup" = "uniqueID")) |> 
  dplyr::rename(`AltTissueID`= dummyID ) |> 
  mutate(pup_SNPed = if_else(is.na(pup_SNPed), FALSE, pup_SNPed))

all_duplicated <- \(x){
  duplicated(x) | duplicated(x, fromLast = TRUE)
}

tissue_options %>%
  mutate(AltTissueID = ifelse(AltTissueID == PupTissueID, NA, AltTissueID)) %>%
  group_by(uniqueID_pup) %>%
  arrange(AltTissueID) %>%
  mutate(numbering = paste0("AltTissueID_", row_number())) %>%
  pivot_wider(id_cols = c(PupTissueID:Beach, Picogreen_con, PlateNumber, PlateLocation),
              names_from = "numbering",
              values_from = "AltTissueID") |> 
  dplyr::select(uniqueID_pup:Beach, everything()) |> 
  arrange(PupBirthyear)  |> 
  datatable(extensions = 'Buttons',
            options = list(dom = 'Blfrtip',
                           buttons = c('copy', 'csv'),
                           pageLength = 10), 
            rownames = FALSE, 
            class = 'cell-border stripe') |>  
  formatStyle(columns = 1:10,
              fontSize = '75%',
              lineHeight = "1em")
```
<br>

Summarize the sample selection by strata (year, recruitment success, SNP availability).

```{r}
#| figure-width: 13
#| fig-height: 6.5
selection_summary <- data_selection |> 
  group_by(PupBirthyear, status, pup_SNPed) |> 
  summarise(n = n(),
            y = y[[1]]) |> 
  mutate(fitness_class = if_else(status %in% c("BeachDead", "NonRecruited"),
                                 "NonRecruited", status))

lab_c <- \(clr, lab){glue::glue("<span style='color:{clr}'>{lab}</span> ")}

data_selection |> 
ggplot(aes(x = PupBirthyear)) +
  geom_rect(aes(xmin = PupBirthyear - .475,
                xmax = PupBirthyear + .475,
                ymin = y - .475,
                ymax = y + .475,
                fill = status),
           color = "transparent") +
  geom_text(data = selection_summary |>
              group_by(status, pup_SNPed) |> 
              summarise(n = sum(n),
                        y = y[[1]]) |> 
              ungroup() ,
            aes(x = 1974, y = y, label = n, color = status),
            family = fnt_sans) +
  geom_text(data = data_selection |> 
              filter(!is.na(Picogreen_con)),
            aes(y = y, label = glue::glue("({PlateNumber}/{PlateLocation}) {round(Picogreen_con)}")),
            angle = 90, color = "white", family = "Josefin sans", fontface = "bold", size = 3) +
  facet_grid(pup_SNPed ~ ., 
             labeller = label_both,
             switch = "y") +
  scale_fill_manual(values = clrs, guide = "none") +
  scale_color_manual(values = clrs, guide = "none") +
  scale_y_continuous(breaks = c(-1.5, -.5, .5),
                     labels = map2_chr(rev(clrs), rev(names(clrs)), lab_c)) +
  coord_cartesian(ylim = c(-2, 1)) +
  theme(axis.text.y = ggtext::element_markdown(angle = 90, hjust = .5),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        strip.placement = "outside")
```

Format and export the sample selection to [`tsv` file](https://github.com/k-hench/arcgaz_gl/blob/master/data/samples_temporal.tsv).

```{r}
tissue_options %>%
  mutate(AltTissueID = ifelse(AltTissueID == PupTissueID, NA, AltTissueID)) %>%
  group_by(uniqueID_pup) %>%
  arrange(AltTissueID) %>%
  mutate(numbering = paste0("AltTissueID_", row_number())) %>%
  pivot_wider(id_cols =  c(PupTissueID:fitness_class, Picogreen_con, PlateNumber, PlateLocation),
              names_from = "numbering",
              values_from = "AltTissueID") |> 
  arrange(-PupBirthyear, uniqueID_pup) |> 
  dplyr::select(unique_id = uniqueID_pup,
                birth_year = PupBirthyear,
                status, fitness_class,
                sex = PupSex,
                SNPed = pup_SNPed,
                beach = Beach,
                plate_number = PlateNumber,
                plate_location = PlateLocation,
                tissue_id = PupTissueID,
                tissue_id_alt1 = AltTissueID_1) |> 
  write_tsv(here("data", "samples_temporal.tsv"))
```

